---
title: "CellTopic analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CellTopic analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  results = "hold",
  comment = "",
  warning = FALSE
)
```

## SpaTopic

### Load data

```{r, message=FALSE}
library(Seurat)
library(SpaTopic)
library(ggplot2)
library(ggforce)
library(cowplot)
library(tidyverse)
library(pheatmap)
my_colors <- c("#66b803", "#E5AA9B", "#FABD3E", "#2B8CBE", "#DE77AE", "#9970AB", "gray", "#D5E4A2", "#71D0F5", "#B1746F", "#ADE2D0", "#20DE8BFF", "#CCDE8BFF", "#FFDE8BFF", "#FFA88BFF", "#FF6A8BFF")
load("vignette_data/HCC_1L_obj.rda")
load("vignette_data/HCC_1L_spot_celltype.rda")
```

The "HCC_1L_obj.rda" comes from <https://www.science.org/doi/10.1126/sciadv.abg3750>.

### cluster

Here we use the clustering method 'Seurat', parameters are as follows, the data has been processed by `SCTransform`.

```{r, message = FALSE}
HCC_1L_obj <- RunPCA(HCC_1L_obj, assay = "SCT", verbose = FALSE)
HCC_1L_obj <- FindNeighbors(HCC_1L_obj, reduction = "pca", dims = 1:30)
HCC_1L_obj <- FindClusters(HCC_1L_obj, resolution = 1.5)
HCC_1L_obj <- RunUMAP(HCC_1L_obj, reduction = "pca", dims = 1:30)
```

### CellTopic

```{r, message = FALSE, out.width = '50%', fig.show = "hold"}
HCC_1L_spot_clusters <- HCC_1L_obj@meta.data["seurat_clusters"]
result_list <- CellTopic(
  HCC_1L_spot_celltype,
  HCC_1L_spot_clusters,
  cluster = "seurat_clusters",
  num_topics = 10,
  percent = 0.7
)
HCC_1L_obj <- AddMetaData(HCC_1L_obj, result_list[["CellTopic"]])
SpatialDimPlot(HCC_1L_obj, group.by = "CellTopic", image.alpha = 0, pt.size.factor = 2.2) + scale_fill_manual(values = my_colors)
SpatialDimPlot(HCC_1L_obj, alpha = 0) + theme(legend.position = "none")
```

## Special spatial domain

SpaTopic can combine cell types and spatial location of CellTopics to analyze certain specialized spatial tissue structures in cancer, such as Tertiary Lymphoid Structures (TLSs) and tumor boundary.

We can see that the slice is divided into 15 CellTopics, of which CellTopics 1, 2, 4, and 5 belong to the cancer regions, and their main cell types are epithelial cells.

```{r, fig.height = 12, fig.width=6}
CellTopic_plot(HCC_1L_obj, 
               celltype_topic = result_list[["celltype_topic"]], 
               celltopic = paste0("CellTopic", c(1,2,4,5)),
               cols.highlight = c("#DE2D26", "#FEE0D2"),
               cols.celltype = my_colors, 
               pt.size.factor = 2.2)
```

CellTopic8 mainly represents the tumor boundary, and its major cell types include fibroblast cells, pDC cells and B cells.

```{r, fig.height = 6, fig.width = 3}
CellTopic_plot(HCC_1L_obj, 
               celltype_topic = result_list[["celltype_topic"]], 
               celltopic = "CellTopic8", 
               cols.highlight = c("#DE2D26", "#FEE0D2"),
               cols.celltype = my_colors,
               pt.size.factor = 2.2)
```

Additionally, CellTopic 7 is found in the paraneoplastic region, predominantly consisting of plasmacytoid dendritic cells (pDCs), B cells, and myeloid cells.

On the right, we observe the scoring for spots using the "[TLS50](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC8683021/)" gene set, which is considered to be the gene signature for the identification of TLSs.

```{r, fig.height = 6, fig.width=5, out.width = '50%', fig.show = "hold"}
CellTopic_plot(HCC_1L_obj, 
               celltype_topic = result_list[["celltype_topic"]], 
               celltopic = "CellTopic7", 
               cols.highlight = c("#DE2D26", "#FEE0D2"),
               cols.celltype = my_colors, 
               highlight.circle = "isTLS",
               pt.size.factor = 2.2)
SpatialFeaturePlot(HCC_1L_obj, features = "scores", pt.size.factor = 2.2, image.alpha = 0) + ggtitle("TLS50")
```

## Meta analysis

### Load data

Presented here is a meta-analysis of the CellTopic-celltypes data derived from four primary colorectal cancer cases as well as two corresponding metastatic liver cancer instances.

```{r}
load("vignette_data/Meta_CellTopic.rda")
Meta_CellTopic[1:5,1:5]
```

Initially, we calculate the correlation among the CellTopics. Based on the complete clustering outcomes, we identify CellTopics that comprise similar cell types and subsequently define these clusters as MetaTopics.

```{r, fig.height = 8, fig.width=8}
corr <- round(cor(Meta_CellTopic), 3)
p <- pheatmap(corr,
         border_color = "black",
         scale = "none", 
         cluster_rows = T,
         cluster_cols = T,
         legend = TRUE,
         show_rownames = T,
         show_colnames = T,
         fontsize = 8, 
         cutree_rows = 7,
         cutree_cols = 7
)
```

Utilizing the Wilcoxon test, we discerned the cell types that exhibit high expression for each MetaTopic. The heatmap below displays the mean expression of each cell type within each MetaTopic, with the p-values indicating the significance of the differences in expression levels.

```{r, message=FALSE}
library(reshape2)
row_cluster <- cutree(p$tree_row,k=7) %>% as.data.frame()
colnames(row_cluster) <- "cluster"
row_cluster$cluster <- paste0("M", row_cluster$cluster)
row_cluster$sample <- gsub("_.*", "", rownames(row_cluster))

Meta_CellTopic_cluster <- merge(as.data.frame(t(Meta_CellTopic)), row_cluster, by = 0)
Meta_CellTopic_cluster_long <- melt(Meta_CellTopic_cluster, id.vars = c("cluster", "Row.names", "sample"))
colnames(Meta_CellTopic_cluster_long) <- c("cluster", "CellTopic", "sample", "celltype_min", "value")
Meta_CellTopic_cluster_long$CellTopic <- as.character(Meta_CellTopic_cluster_long$CellTopic)
Meta_CellTopic_cluster_long$celltype_min <- as.character(Meta_CellTopic_cluster_long$celltype_min)
```

```{r, fig.width=5,fig.height=15}
df <- Meta_CellTopic_cluster_long %>% 
  group_by(cluster, celltype_min) %>%
  summarise(mean = mean(value, na.rm=T)) %>%
  spread(celltype_min, mean) %>%
  column_to_rownames(var = "cluster")
df_p <- unique(Meta_CellTopic_cluster_long$celltype_min) %>% as.data.frame() %>% column_to_rownames(var = ".")
ct_n <- 1
for(ct in unique(Meta_CellTopic_cluster_long$celltype_min)){
  all <- Meta_CellTopic_cluster_long %>% subset(celltype_min == ct)
  cls_n <- 1
    for(cls in unique(Meta_CellTopic_cluster_long$cluster)){
      select <- Meta_CellTopic_cluster_long %>% subset(celltype_min == ct & cluster == cls)
      p <- wilcox.test(select$value, all$value, alternative = "greater")
      p <- p$p.value
      df_p[ct_n,cls_n] <- p
      cls_n <- cls_n + 1
    }
  ct_n <- ct_n + 1
}
colnames(df_p) <- unique(Meta_CellTopic_cluster_long$cluster)

df_pp <- df_p %>% t()
df_pp_cluster <- apply(df_pp, 2, function(x){ifelse(min(x) <= 0.05, return(which.min(x)), return("none"))}) %>% 
  as.data.frame()
df_pp_cluster_name <- rownames(df_pp_cluster)[order(df_pp_cluster$.)]
df_pp <- df_pp[,df_pp_cluster_name]
df_pp_cluster <- apply(df_pp, 2, function(x){ifelse(min(x) <= 0.05, return(which.min(x)), return("none"))}) %>% 
  as.data.frame()
df_pp_cluster$. <- as.character(df_pp_cluster$.)
colnames(df_pp_cluster) <- "high in MetaTopics"
mycolor <- c("#f94144", "#f8961e", "#f7f31c", "#77bb43", "#429f9c", "#3574b8", "#bb3892", "grey")
names(mycolor) <- c("1", "2", "3", "4", "5", "6", "7", "none")
df[,df_pp_cluster_name] %>% t() %>%
  pheatmap(border_color = "black", scale = "none", legend = TRUE, 
    cellwidth = 15, cellheight = 15,
    cluster_rows = F, cluster_cols = F, 
    show_rownames = T, show_colnames = T,
    fontsize = 8, fontsize_number = 9, number_color = "black", 
    display_numbers = t(matrix(ifelse(df_pp <= 0.05, ifelse(df_pp <= 0.01, "***", "*"), ""), nrow(df_pp))),
    gaps_row = c(9,16,21,25,34,45,56),
    annotation_row = df_pp_cluster,
    annotation_colors = list(`high in MetaTopics` = mycolor),
    angle_col = 90
  )
```

## Cell-Cell communication

We can delve deeper into the complexities of the tumor microenvironment by analyzing the communication among different CellTopics. Employ the [CellChat](https://github.com/jinworks/CellChat) tool to elucidate the dynamics of cell communication across spatial domains.

### load data

```{r}
data("spot_celltype")
data("spot_clusters")
load("vignette_data/PDAC_obj.rda")
result_list <- CellTopic(spot_celltype, spot_clusters, cluster = "spatial.cluster", num_topics = 13)
PDAC_obj <- AddMetaData(PDAC_obj, result_list[["CellTopic"]])
```

```{r, message = FALSE}
library(CellChat)
# Create cellchat object
data.input <- GetAssayData(PDAC_obj, assay = "Spatial", slot = "data")
identity = data.frame(group =PDAC_obj$CellTopic, row.names = names(PDAC_obj$CellTopic))
cellchat <- createCellChat(object = data.input, meta = identity, group.by = "group")
# Selective ligand receptor database
cellchat@DB <- CellChatDB.human
cellchat <- subsetData(cellchat)
# Calculation
cellchat <- identifyOverExpressedGenes(cellchat)
cellchat <- identifyOverExpressedInteractions(cellchat)
cellchat <- projectData(cellchat, PPI.human)
#cellchat <- filterCommunication(cellchat, min.cells = 10) # Do not filter spot
cellchat <- computeCommunProb(cellchat, raw.use = FALSE, population.size = TRUE)
cellchat <- computeCommunProbPathway(cellchat)
df.net <- subsetCommunication(cellchat)
```

Communication between different spatial domains from the perspective of pathway.

```{r, fig.height = 10, fig.width = 4}
df.netp <- subsetCommunication(cellchat, slot.name = "netP")
df.netp$sTot <- paste0(df.netp$source, " -> ", df.netp$target)
df.netp <- subset(df.netp, sTot %in% c("CellTopic1 -> CellTopic1","CellTopic1 -> CellTopic2","CellTopic1 -> CellTopic3","CellTopic1 -> CellTopic4","CellTopic2 -> CellTopic1","CellTopic3 -> CellTopic1","CellTopic4 -> CellTopic1","CellTopic2 -> CellTopic4","CellTopic4 -> CellTopic2","CellTopic2 -> CellTopic2","CellTopic3 -> CellTopic3","CellTopic4 -> CellTopic4"))
df.netp$sTot <- factor(df.netp$sTot,levels = c("CellTopic2 -> CellTopic4","CellTopic4 -> CellTopic2","CellTopic1 -> CellTopic2","CellTopic2 -> CellTopic1","CellTopic1 -> CellTopic3","CellTopic3 -> CellTopic1","CellTopic1 -> CellTopic4","CellTopic4 -> CellTopic1","CellTopic1 -> CellTopic1","CellTopic2 -> CellTopic2","CellTopic3 -> CellTopic3","CellTopic4 -> CellTopic4"))
df.netp <- rbind(df.netp, data.frame(
    source = c(NA,NA,NA), 
    target = c(NA,NA,NA), 
    pathway_name = c("VTN","VTN","VTN"), # It's just for showing the parts without communication, "VTN" has no particularity.
    prob = c(0,0,0), 
    pval = c(NA,NA,NA), 
    sTot = c("CellTopic1 -> CellTopic2","CellTopic2 -> CellTopic1","CellTopic3 -> CellTopic1")))
library('RColorBrewer')
ggplot(df.netp, aes(x=sTot, y=pathway_name)) + 
    geom_point(aes(size=pval, color=log10(prob))) + 
    scale_size(range = c(4, 2), breaks = c(0.01,max(df.netp$pval[!is.na(df.netp$pval)])), labels = c("p<0.01","0.01<p<0.05")) + 
    scale_colour_distiller(palette="Spectral", labels = c("low","high"), breaks = c(log10(min(df.netp$prob[df.netp$prob!=0])),log10(max(df.netp$prob)))) +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    axis.title.x = element_blank(),
    axis.title.y = element_blank()) +
    labs(size = "p-value", color = "Commun.prob")
```
