---
title: "Quick start"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "",
  warning = FALSE
)
```

```{r}
library(SpaTopic)
library(ggplot2)
my_colors <- c("#66b803", "#E5AA9B", "#FABD3E", "#2B8CBE", "#DE77AE", "#9970AB", "gray", "#D5E4A2", "#71D0F5", "#B1746F", "#ADE2D0", "#20DE8BFF", "#CCDE8BFF", "#FFDE8BFF", "#FFA88BFF", "#FF6A8BFF")
```

## Load data

SpaTopic requires two inputs:

-   `spot_celltype`: The result of the deconvolution of single-cell and spatial transcriptome data, or a matrix with rows as spots and columns as cell types.\*

-   `spot_clusters`: A dataframe with spot domain information.\*

And they should be like

```{r}
data("spot_celltype")
data("spot_clusters")
head(spot_celltype)
```

```{r}
head(spot_clusters)
```

\*The data used here is from [CARD](https://github.com/YMa-lab/CARD) (`spot_celltype`) and [BayesSpace](https://github.com/edward130603/BayesSpace) (`spot_clusters`).

### About data

The corresponding Seurat object data can also be obtained from [GEO](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE111672). (Data used here is processed by SCTransform) The `spot_celltype` and `spot_clusters` is based on it.

```{r, message = FALSE, out.width = '50%', fig.show = "hold", fig.height = 4, fig.width = 5}
load("vignette_data/PDAC_obj.rda")
PDAC_obj <- AddMetaData(PDAC_obj, spot_clusters["spatial.cluster"])
DimPlot(PDAC_obj, group.by = 'spatial.cluster', reduction = "umap",cols = my_colors)
SpatialDimPlot(PDAC_obj, label = FALSE, group.by = "spatial.cluster", pt.size.factor = 8) + scale_fill_manual(values = my_colors) 
```

## Simply usage

Enter the deconvolution result `spot_celltype` and the grouping information of the spot `spot_clusters` to get an `result_list` :

-   `CellTopic` is a dataframe which can be add to a Seurat object.

-   `domain_topic` is a dataframe, row is CellTopics and col is domains.

-   `celltype_topic` is a dataframe, row is cell types and col is CellTopics.

-   `Cell_topic` is a vector of which topic can be chosen in each CellTopic.

For subsequent analyses, `CellTopic` and `celltype_topic` are mainly used.

```{r}
result_list <- CellTopic(
  spot_celltype,
  spot_clusters,
  cluster = "spatial.cluster",
  num_topics = 13
)
```

## Plot

You can add the results of `CellTopic` to meta.data of Seurat object through `AddMetaData` of Seurat, and display the spatial position of all CellTopics through `SpatialDimPlot` of Seurat.

```{r, fig.height = 4, fig.width = 5}
PDAC_obj <- AddMetaData(PDAC_obj, result_list[["CellTopic"]])
SpatialDimPlot(PDAC_obj, group.by = "CellTopic",pt.size.factor = 8)
```

And `CellTopic_plot` can be used to display the spatial location and cell type of one or more CellTopics at the same time.

Please use the Seurat object with CellTopic added as meta.data and the `celltype_topic` dataframe in the result list of `FindCellTopic` (or `CellTopic`) function as input.

```{r, fig.height = 12, fig.width = 6}
CellTopic_plot(PDAC_obj, 
               celltype_topic = result_list[["celltype_topic"]], 
               cols.highlight = c("#DE2D26", "#FEE0D2"),
               cols.celltype = my_colors, 
               pt.size.factor = 8)
```

## Optional parameters

```{r}
result_list <- CellTopic(
  spot_celltype,
  spot_clusters,
  cluster = "spatial.cluster",
  num_topics = 13,
  percent = 0.7
)
```

We recommend that the value of the `num_topics` is generally slightly larger than the group number of spots.

The larger the `percent`, the more dependent the results of deconvolution. We recommend using a value of 0.5-0.9.

### Another strategy

If you want to focus less on percentages and more on the information carried by a single topic, you can use `Binarization = TRUE` .

```{r, fig.height = 12, fig.width = 6}
result_list <- CellTopic(
  spot_celltype,
  spot_clusters,
  cluster = "spatial.cluster",
  num_topics = 13,
  Binarization = TRUE
)
PDAC_obj <- AddMetaData(PDAC_obj, result_list[["CellTopic"]])
CellTopic_plot(PDAC_obj, 
               celltype_topic = result_list[["celltype_topic"]], 
               cols.highlight = c("#DE2D26", "#FEE0D2"),
               cols.celltype = my_colors, 
               pt.size.factor = 8)
```

### Meta Topic

If you want to explore different spatial domains based on the cell type of the CellTopic, we provide a simple clustering option, `meta.cell = TRUE`.

```{r}
result_list <- CellTopic(
  spot_celltype,
  spot_clusters,
  cluster = "spatial.cluster",
  num_topics = 13,
  percent = 0.7,
  meta.cell = TRUE,
  k = 3
)
head(result_list[["MetaTopic"]])
```

Parameter k is a integer number of MetaTopics. Hierarchical clustering algorithm is used by default
