---
title: "Spatial expansion"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatial expansion}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  results = "hold",
  comment = "",
  warning = FALSE
)
```

```{r setup, , message=FALSE}
library(Seurat)
library(ggplot2)
library(SpaTopic)
```

## Load data

```{r}
my_colors <- c("#66b803", "#E5AA9B", "#FABD3E", "#2B8CBE", "#DE77AE", "#9970AB", "gray", "#D5E4A2", "#71D0F5", "#B1746F", "#ADE2D0", "#20DE8BFF", "#CCDE8BFF", "#FFDE8BFF", "#FFA88BFF", "#FF6A8BFF")
load("vignette_data/HCC_1L_obj.rda")
load("vignette_data/HCC_1L_spot_celltype.rda")
```

## SpaTopic

We first use SpaTopic to get the CellTopic of the spatial domain.

```{r, message=FALSE}
HCC_1L_obj <- RunPCA(HCC_1L_obj, assay = "SCT", verbose = FALSE)
HCC_1L_obj <- FindNeighbors(HCC_1L_obj, reduction = "pca", dims = 1:30)
HCC_1L_obj <- FindClusters(HCC_1L_obj, resolution = 1.5)
HCC_1L_obj <- RunUMAP(HCC_1L_obj, reduction = "pca", dims = 1:30)
HCC_1L_spot_clusters <- HCC_1L_obj@meta.data["seurat_clusters"]
result_list <- CellTopic(
  HCC_1L_spot_celltype,
  HCC_1L_spot_clusters,
  cluster = "seurat_clusters",
  num_topics = 10,
  percent = 0.7
)
HCC_1L_obj <- AddMetaData(HCC_1L_obj, result_list[["CellTopic"]])
SpatialDimPlot(HCC_1L_obj, group.by = "CellTopic", image.alpha = 0, pt.size.factor = 2.2) + scale_fill_manual(values = my_colors)
```

## Spatial expansion

`Spatial_expansion` then divides buffers and angles for the specified spatial domain. It is worth noting that this method is not only applicable to SpaTopic, but also to arbitrary spatial partitions with certain continuity.

We divide the buffers and angles of the spatial domain of the tumor boundary by specifying CellTopic column in `Seurat@meta.data`, and CellTopic8 in it.

```{r}
HCC_1L_obj <- Spatial_expansion(HCC_1L_obj,
    CellTopic = c("CellTopic8"),
    meta.col = "CellTopic",
    type = "Hexagon",
    edge.plot = TRUE
)
```

When set `edge.plot = TRUE`, the function draw a reference plot representing the specified spatial domain divide. We can obtain a more appropriate spatial domain divide by modifying the alpha parameter of alpha-shape calculation.

We can check the relevant parameter settings in `Seurat@misc` as a reference for our adjustments

```{r}
HCC_1L_obj@misc$expansion$CellTopic8[1:5]
```

Here we get a more appropriate spatial domain divide by setting `alpha = 280`.

```{r}
HCC_1L_obj <- Spatial_expansion(HCC_1L_obj,
    CellTopic = c("CellTopic8"),
    meta.col = "CellTopic",
    alpha = 280,
    type = "Hexagon",
    edge.plot = TRUE
)
```

We can use `SpatialFeaturePlot` in Seurat to show buffer divide of spatial domain.

```{r}
SpatialFeaturePlot(HCC_1L_obj, features = "expansion_CellTopic8", pt.size.factor = 2.2)
```

On the other hand, you can increase or decrease the width of the buffers by setting `distance` .

```{r}
HCC_1L_obj_1 <- Spatial_expansion(HCC_1L_obj,
    CellTopic = c("CellTopic8"),
    meta.col = "CellTopic",
    distance = 800,
    alpha = 280,
    type = "Hexagon",
    edge.plot = FALSE
)
SpatialFeaturePlot(HCC_1L_obj_1, features = "expansion_CellTopic8", pt.size.factor = 2.2)
```

### Angle

`Spatial_expansion` creates not only buffer information in Seurat\@meta.data, but also angle information.

We regard the geometric center of a spatial domain as the center, and the left side of the spatial domain is the starting point and the ending point of the circle.

```{r}
SpatialFeaturePlot(HCC_1L_obj, features = "angle_CellTopic8", pt.size.factor = 2.2)
```

## Gene plot

After dividing the spatial domain we can use `Expansion_gene_plot` to show gene expression changes in buffers in different angles.

Here we set `angle = 90`, `parts = 2` to get gene expression on both sides of the tumor boundary.

```{r, message=FALSE}
Expansion_gene_plot(HCC_1L_obj,
    CellTopic = c("CellTopic8"),
    gene = c("ACTA2", "C11orf96", "CALD1", "COL1A2", "CSRP1", "CTGF", "FLNA", "IGFBP7", "MYH11", "MYL9", "PPP1R14A", "TAGLN", "TPM2"),
    meta.col = "CellTopic", angle = 90, parts = 2, plot = TRUE, cols = my_colors
)
```

This method was consistent across different angles, as shown when dividing the spatial domain into four sections starting at 50 degrees.

```{r, message=FALSE}
Expansion_gene_plot(HCC_1L_obj,
    CellTopic = c("CellTopic8"),
    gene = c("ACTA2", "C11orf96", "COL1A2", "CSRP1"),
    meta.col = "CellTopic", angle = 50, parts = 4, plot = TRUE, cols = my_colors
)
```

If you need to plot data for further analysis or re-plotting, you can set `plot = FALSE` to obtain the expression of genes in buffers at different angles.
