---
title: "SpaTopic workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SpaTopic workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  results = "hold",
  comment = "",
  warning = FALSE
)
```

## Load data

```{r, message = FALSE}
library(SpaTopic)
library(ggplot2)
library(cowplot)
library(pheatmap)
my_colors <- c("#66b803", "#E5AA9B", "#FABD3E", "#2B8CBE", "#DE77AE", "#9970AB", "gray", "#D5E4A2", "#71D0F5", "#B1746F", "#ADE2D0", "#20DE8BFF", "#CCDE8BFF", "#FFDE8BFF", "#FFA88BFF")
data(spot_celltype)
data(spot_clusters)
load("vignette_data/PDAC_obj.rda")
```

## SpaTopic workflow
The workflow of SpaTopic is detailed below, including Kolmogorov-Smirnov (KS) test, Latent Dirichlet Allocation (LDA), and FindCellTopic.

### KS test

First, we use KS test to calculate the specificity score for cell types across different spot domains.

```{r, fig.height = 3, fig.width = 6}
domain_cellytpe <- ksTest(spot_celltype, spot_clusters, cluster = "spatial.cluster")
pheatmap(domain_cellytpe, cluster_rows = FALSE, cluster_cols = FALSE,
         cellwidth = 20, cellheight = 20)
```

### LDA model

Then we can get the topic that can represent the spatial domain through the LDA model.

```{r}
topic_list <- MatrixFactorization(domain_cellytpe, num_topics = 13)
dt_topic_data <- topic_list[["domain_topic"]]
ct_topic_data <- topic_list[["celltype_topic"]]
```

Show each topic from the LDA model.

```{r}
# plot prepare
PDAC_obj <- AddMetaData(PDAC_obj, spot_clusters["spatial.cluster"])
df <- as.data.frame(t(dt_topic_data[,match(PDAC_obj$spatial.cluster,gsub("spot_domain_", "", colnames(dt_topic_data)))]))
rownames(df) <- colnames(PDAC_obj)
PDAC_obj <- AddMetaData(PDAC_obj, df)
colors_num <- ncol(dt_topic_data)
topics <- colnames(ct_topic_data)
bar_plot_data <- ct_topic_data
bar_plot_data$CellType <- rownames(ct_topic_data)
```

```{r, message = FALSE, fig.height = 30, fig.width = 13}
plot_list <- list()
for (i in seq_along(topics)){
  p <- SpatialPlot(PDAC_obj, group.by = topics[i], pt.size.factor = 8, stroke = NA) + scale_fill_manual(values = rev(c("#CB181D", "#FB6A4A", "#FCAE91", "#FEE5D9"))) + NoAxes() + NoLegend()
  bar_plot <- ggplot(bar_plot_data, aes(x = reorder(CellType, !!sym(topics[i]), decreasing = TRUE), y = .data[[topics[i]]])) +
    geom_bar(aes(fill = CellType), stat='identity', width = 0.7) +
    xlab('CellType') +
    scale_fill_manual(values = my_colors) +
    theme_minimal() + NoGrid() + 
    theme(legend.key.size = unit(0.3, "cm"), 
          legend.position = "none",
          axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0, size = 8))
  plot_list[[i]] <- plot_grid(plotlist = list(p, bar_plot), ncol = 1)
}
plot_grid(plotlist = plot_list, ncol = 3)
```

### FindCellTopic

We assign each spot domain to one or more specific topics (termed 'CellTopics')

```{r, message = FALSE, out.width = '50%', fig.show = "hold", fig.height = 6, fig.width = 5}
result_list <- FindCellTopic(dt_topic_data, ct_topic_data, spot_clusters, cluster = "spatial.cluster", percent = 0.7)
PDAC_obj <- AddMetaData(PDAC_obj, result_list[["CellTopic"]])
SpatialDimPlot(PDAC_obj, label = FALSE, group.by = "CellTopic", pt.size.factor = 8) + scale_fill_manual(values = my_colors) + coord_flip()+ scale_y_reverse()
pheatmap(result_list[["celltype_topic"]], 
               cluster_rows = FALSE, 
               cellwidth = 20, cellheight = 20)
```

When use `Binarization = TRUE`, It returns a result of the same form, but focus less on percentages and more on the information carried by a single topic.

```{r, eval=FALSE}
# Not run, just show
result_list <- FindCellTopic(dt_topic_data, ct_topic_data, spot_clusters, cluster = "spatial.cluster",Binarization = TRUE)
```

### Plot

`CellTopic_plot` can be used to display the spatial location and cell types of one or more CellTopics at the same time.

```{r, fig.height = 12, fig.width = 5}
CellTopic_plot(PDAC_obj, 
               celltype_topic = result_list[["celltype_topic"]], 
               cols.highlight = c("#DE2D26", "#FEE0D2"),
               cols.celltype = my_colors, 
               pt.size.factor = 8)
```
